Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-04 manuscripts/manuscript.Rmd
edits as sent to Marc

<!-- __We don't have the data where we need it__ -->
<!-- What do we call this?  Extrapolation uncertainty?  Pathological Uncertainty? -->

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated, and models may be abused [@Mangel2001].
Extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013].

<!--This difficult position of having neither the true model nor data that
covers the full range of possible states is unfortunately the rule more
than the exception. The potential concern of tipping points in ecological
dynamics [@Scheffer2001; @Polasky2011] reflects these concerns -- as
either knowledge of the true model or more complete sampling of the
state space would make it easy to identify if a tipping point existed.
If we do not know but cannot rule out such a possibility, then we
face decision-making under this 
-->

@@

We illustrate how a Stochastic Dynamic Programming (SDP) algorithm
[@Mangel1988; @Marescot2013] can be driven by the predictions from a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
This provides two distinct advantages compared with contemporary
approaches.  First, using a BNP sidesteps the need for an accurate
model-based description of the system dynamics.  Second, the BNP can
better reflect uncertainty that arises when extrapolating a model outside
of the data on which it was fit.  We illustrate that when the correct
model is not known, this latter feature is crucial to providing a robust
decision-theoretic approach in face of substantial structural uncertainty.

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)), t)  \right] \}$$
\end{equation}

@@

We consider the fase where we have some previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm obs}(t), X_{\textrm obs}(t+1)$ for $t \in [1,2, ..., T]$

Under the GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau in [T+1, T+2, ..., T+n]$, is simply given by a multivariate normal ($\mathcal{N}$):

\begin{equation}
g(X_p|X_{\textrm obs}(t), X_{\textrm obs}(t+1)) \sim \mathcal{N}(E,C)
\end{equation}

@@

\begin{equation}
E = K(X_{\textrm p}(\tau), X_{\textrm obs}(t)) \left(K(X_{\textrm obs}(t),X_{\textrm obs}(t)) - \sigma \mathbb{I} \right)^{-1} X_{\textrm obs}(t+1)
\end{equation}

\begin{equation}
C = K(X_{\textrm p}(\tau), X_{\textrm p}(\tau)) - K(X_{\textrm p}(\tau), X_{\textrm obs}(t)) \left(K(X_{\textrm obs}(t),X_{\textrm obs}(t)) - \sigma \mathbb{I} \right)^{-1} K(X_{\textrm obs}(t), X_{\textrm p}(\tau))
\end{equation}

@@

These results illustrate that model-choice approaches would be positively
misleading -- supporting simpler models that cannot express tipping point
dynamics merely on account of them being similar.  As the data shown
comes only from the basin of attraction near the unfished equilibrium,
near which all of the models are approximately linear and approximately
identical.

Model selection criteria trade off model complexity and fit to the data. 
When the data come from a limited region of state-space -- as is necessarily 
the case whenever there is a potential concern about tipping point dynamics --
simpler models therefore fit just as well and tend to be selected more often than complex 
ones.  This approach would be appropriate when the dynamics can be expected 
to remain in the region of the training data; for instance, if we only 
considered the forecasting accuracy of the unfished population dynamics under
each model.

In contrast, the decision-maker's problem of setting appropriate harvest levels
cannot exclude regions of state-space outside the observed range when integrating
over all possible decisions to find the optimal choice.  Such problems are not 
constrained to fisheries management but ubiquitous across ecological decision-making
and conservation where the greatest concerns involve entering previously unobserved
regions of state-space -- whether that is the collapse of a fishery, the spread
of an invasive, or the loss of habitat.

@@

The parametric models perform worst when they propose a management strategy
outside the range of the observed data. The non-parametric Bayesian approach, 
in contrast, allows a predictive model that expresses a great deal of uncertainty
about the probable dynamics outside the observed range, while retaining very
good predictive accuracy in the range observed.  The management policy 
dictated by the GP balances this uncertainty against
the immediate value of the harvest, and acts to stabilize the population 
dynamics in a region of state space in which the predictions can be 
reliably reflected by the data.

>>>>

<!-- Point is: __We don't have the data where we need it__ -->
<!-- What do we call this?  Extrapolation uncertainty?  Pathological Uncertainty? -->

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated, and models may be abused [@Mangel2001].
Extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the space of 
all possible actions involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013], hence estimates of the uncertainty can be as
important as estimates of the average or expected outcome.

Concerns about tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because they arise from this very
situation.  In such cases, precise models are not available and data is
limited or biased to a particular region of state space, such as high
stock levels or an otherwise desirable state for the ecosystem. If such
models or data were available, one would know just how far a system could
be pushed before going over the tipping point, and management would be
simple.  The fact that this concern is frequently discussed highlights
how often we have only limited data and inperfect models. Moreover,
the case of tipping points highlights how even small errors can have
very large consequences, as we shall illustrate later.

@@

We illustrate how a Stochastic Dynamic Programming (SDP) algorithm
[@Mangel1988; @Marescot2013] can be driven by the predictions from a
Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
This provides two distinct advantages compared with contemporary
approaches.  First, using a BNP sidesteps the need for an accurate
model-based description of the system dynamics.  Second, the BNP can
better reflect uncertainty that arises when extrapolating a model outside
of the data on which it was fit.  We illustrate that when the correct
model is not known, this latter feature is crucial to providing a robust
decision-theoretic approach in face of substantial structural uncertainty.
We use _robust_ to distinguish approaches that provide nearly optimal solutions
(as measured by the chosen value function) without being sensitive to the
choice of the (unknown) underlying model.

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)), t)  \right] \}
\end{equation}

@@

We consider the fase where we have some previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t \in [1,2, ..., T]$ up through observed time $T$.  Under the GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is simply given by a multivariate normal ($\mathcal{N}$):

\begin{equation}
g(X_p|X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)) \sim \mathcal{N}(E,C)
\end{equation}

@@

\begin{equation}
E = K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} X_{\textrm{obs}}(t+1)
\end{equation}

\begin{equation}
C = K(X_{\textrm p}(\tau), X_{\textrm p}(\tau)) - K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} K(X_{\textrm{obs}}(t), X_{\textrm p}(\tau))
\end{equation}

@@

These results illustrate that model-choice approaches would be positively
misleading -- supporting simpler models that cannot express tipping point
dynamics merely on account of them being similar.  The data shown
comes only from the basin of attraction near the unfished equilibrium,
near which all of the models are approximately linear and approximately
identical; so it is intuitive that all criteria choose the simplest model.
In a complex world, the result is suboptimal. But in a world that might
contain tipping points, the result is disastrous.

Many managers both in fisheries and beyond face a similar situation: they
have not observed the population dynamics at all possible densities.  A
lack of comprehensive data at all population sizes, combined with our obvious
inability to formulate accurate population models for low population sizes
in the absence of any data, makes this situation the rule more than the exception.
Relying on parametric models and model choice processes that favor simplicity
ignores this basic fact. Acknowledging this, Carl Walters [e.g. @Walters1978] has long observed that if we began
by fishing any newly exploited population down to very low levels and
then let it recover, we would be much better at estimating population
dynamics and thus predicting the optimal harvest levels.  While certainly true,
this presents a rather risky policy in the face of potential tipping points. 
The GPDP offers an alternative way to acknowledge this uncertainty.

@@

The parametric models perform worst when they propose a management strategy
outside the range of the observed data. The non-parametric approach of the GPDP, 
in contrast, allows a predictive model that expresses a great deal of uncertainty
about the probable dynamics _outside_ the observed range, while retaining very
good predictive accuracy _inside_ the range observed.  The management policy 
dictated by the GPDP balances this uncertainty against
the immediate value of the harvest, and acts to stabilize the population 
dynamics in a region of state space in which the predictions can be 
reliably reflected by the data.

Such problems are not 
constrained to fisheries management but ubiquitous across ecological decision-making
and conservation where the greatest concerns involve entering previously unobserved
regions of state-space -- whether that is the collapse of a fishery, the spread
of an invasive, or the loss of habitat.
